Categorical and gradient aspects of wordlikeness\thanks{Thanks to Gene Buckley, Charles Yang, and audiences at NELS 43 for helpful comments.}

### 2012-11-11 2.tex
Completed writing on 2-4!

FIXME
%\begin{quote}
%Hence, a real solution to the problem of ``admissibility'' will not simply define a tripartite categorization of occurring, accidental gap, and inadmissible, but will define the `degree of admissibility' of each potential lexical matrix in such a way as to\ldots{}make numerous other distinctions of this sort (\emph{SPE}:416--417)
%\end{quote}
%
%\noindent
%The position that the well-formedness of nonce words is consistent the view of syntactic grammaticality taken in \emph{LSLT} and \emph{Aspects} \citep{LSLT,ASPECTS}, where it is claimed that different syntactic violations result in different degrees of ungrammaticality.
%
%Most recent discussions of the ``problem of admissibility'' focus on this cline of wellformedness as it is evidenced in wordlikeness tasks.
%
%\begin{quote}
%When native speakers are asked to judge made-up (nonce) words, their intuitions are rarely all-or-nothing. In the usual case, novel items fall along a gradient cline of acceptability. \citep[][9]{Albright2009a}
%
%In the particular domain of phonotactics gradient intuitions are pervasive: they have been found in every experiment that allowed participants to rate forms on a scale.
%\citep[][382]{Hayes2008a}
%
%\ldots{}when judgements are elicited in a controlled fashion from speakers, they always emerge as gradient, including all intermediate values. \citep[371]{Shademan2006} 
%\end{quote}
%
%\begin{quote}
%\ldots{}A defect of current grammatical acounts of phonotactics is that they render simple up-or-down decisions concerning well-formedness and cannot account for gradient judgements. \citep[371]{Shademan2006}
%\end{quote}

\section{What some wellformedness judgements might not be}

@@

%The gradedness of a category cannot be observed. Scientists choose the granularity at which they observe. For example, a linguist may choose to record speakers' judgements on a 7-point scale, for example, or to use a forced binary choice between two minimally different items. The level of granularity at which the observations are made have consequences for how statistical inference over these observations should be performed, or for the theory of the \emph{task model}, the mechanism by which speakers perform the task, but it does not have any consequences for the \emph{theory} of phnotactics.

%\begin{quote}
%No integer seems to sit on the fence, undecided as to whether it is quite even, or perhaps a bit odd. No odd number seems odder than any other odd number. \citep[274]{Armstrong1983}
%\end{quote}

%\begin{quote}
%Some have responded to these findings very consistently, by asserting that the experimental findings are to be interpreted as before: that, psychologically speaking, odd numbers as well as birds and vegetables are graded concepts\ldots{} We reject this conclusion just because we could not explain how a person could compute with integers who believed that 7 was odder than 23. We assert confidently that the facts about subjects being able to compute and about their being able to give the definition of odd number, etc., are the more important, highly entrenched, facts we want to preserve and explain\ldots{} we ourselves are prepared to give up the seeming fact that some odd numbers appear, as shown by their behavior in certain experimental paradigms, to be odder than others\ldots{}we do not give it up by saying that it was no fact; rather, by saying it must have been a fact about something other than the structure of concepts. \citep[284]{Armstrong1983}
%\end{quote}

%\begin{quote}
%\ldots{}we hold that \emph{fruit} and \emph{odd number} have different structures, and yet we obtain the same experimental outcome for both. But if the same result is achieved regardless of the concept struture, then the experimental design is not pertinent to the determination of concept structure. \citep[284--5]{Armstrong1983}
%\end{quote}

@@

There have been no serious attempts to evaluate categorical and gradient models of wordlikeness on an equal footing. This study represents a first attempt to fill this gap. It is not that categorical models have been ignored by the literature on wordlikeness modeling, but rather that they have not been compared. For instance, \citet{Hayes2008a}, who compare their gradient model of wordlikeness against a set of phonotactic constraints proposed by \citet{Clements1983}, transform the \citeauthor{Clements1983} constraints, many of which are exceptionless, into probabilities. While this is consistent with their claim that ``the ability to model gradient intuitions to be an important criterion for evaluating phonotactic models'' \citep[382]{Hayes2008a}, such a principle precludes any attempt to confirm or falsify the hypothesis that underlies it.

It is unknown whether the intermediate ratings in gradient wordlikeness tasks are reliably predicted by the computational models that have been proposed. If some model poorly accounts for these intermediate ratings, it may be the case that the model is improperly specified, or it may indicate that a considerable amount of variance in wordlikeness ratings is not a product of the phonotactic system per se.

Imagine, alternatively, that wordlikeness judgements could be effectively modeled with a gross contrast between possible and impossible words.

%[st] [bl]

@@

This evaluation uses a large sample of three previously published studies on English wordlikeness comprising 125 subjects and 187 items. Two criteria were used to select these three studies. First, it was required that the stimuli be presented aurally so as to eliminate any possibility of orthographic effects. Secondly, the data must be sufficiently ``phonotactically diverse'': that is, it must include both items like \emph{blick} and \emph{bnick}. This excludes studies like that of \citet{Bailey2001}, in which few if any items contain gross phonotactic violations of the type represented by \emph{bnick}. The data is summarized in Table \ref{counts}.

@@

\citet{Albright2007} administers a wordlikeness task in which 68 adult speakers rate 40 monosyllabic nonce words, presented aurally, on a 7-point Likert scale with endpoints labeled  ``completely impossible as an English word'' and ``would make a fine English word''. \citealt{Albright2007}'s study study is primarily concerned with the effects of different onset types (e.g., well-formed /bl/, marginal /bw/, unattested /bn, bd, bz/), and there is less diversity among the choice of rimes, none of which are obviously ill-formed.

@@

\citet{Scholes1966} conducts several wordlikeness tasks with 7th-grade students (approximately 12--13 years of age). The data used here is his experiment 5, in which 33 speakers provide a ``yes'' or ``no'' as to whether each of the 63 items, presented aurally, are ``likely to be usable as a word of English''. Like the study by \citet{Albright2007}, the focus was on onset well-formedness and there is minimal diversity in rime type. Two items, [klʌŋ] \emph{clung} and [bɹʌŋ] \emph{brung} (a dialectical past participle of \emph{bring}), are excluded here as actual words of English. \citet{Albright2009a} and \citet{Hayes2008a} also use this data for the purposes of model evaluation; following \citet{Frisch2000}, they use the proportion of ``yes'' responses for each item so as to derive a continuous measure of well-formedness.

@@

Models are evaluated by comparing their scores to the average rating of each word using four correlation statistics. \citet{Hayes2008a} evaluate their model using the Pearson $r$, a parametric measure of correlation. \citet{Stevens1946} argues that statistics of this type are inappropriate for analysis of Likert scale data, like those used by \citet{Albright2009a} and \citet{Albright2003b}. The reason is that this makes a \emph{linearity assumption}. That is, it assumes that nonce words rated ``1'' and  ``3'', for instance, are just as different as those ``4'' and a ``6''. A weaker assumption, more appropriate for Likert scale data, is the ``monotonicity assumption'': that ``1'' is less English-like than ``3'', which is less English-like than ``4'', and so on. However, it also has been claimed that $r$ is particularly robust to violations of the linearity assumption \citep[e.g.,][]{Havlicek1976}. Pearson $r$ is reported here, but no stance is taken on its appropriateness for this data.
%\citet{Jamieson2004}
\citeauthor{Hayes2008a} also report Spearman $\rho$; this statistic requires only the weaker assumption of monotonicity, but it is difficult to give a simple interpretation to the coefficient. Much easier to interpret are two non-parametric statistics, the Goodman-Kruskal $\gamma$ and the Kendall $\tau_b$ \citep{Noether1981}; a variant of the latter is used by \citet{Albright2009a}. These statistics are computed by comparing every model score/wordlikeness rating pair to every other: a comparison is counted as \emph{concordant} if the greater of the two model scores is the one associated with the greater of the two wordlikeness ratings (that is, the model ranks these two nonce words the in accordance with speakers' ratings), and as \emph{discordant} otherwise. $\gamma$ and $\tau_b$ differ in the treatment of ``ties'', pairs where either the model score or wordlikeness rating are identical. For $\gamma$, ties are ignored, and the coefficient is

\begin{unlabeledexample}
$\gamma = \displaystyle\frac{c - d}{c + d}$
\end{unlabeledexample}

\noindent
where $c$ and $d$ represent the number of concordant and discordant pairs, respectively. The $\tau_b$ statistic uses a similar formula, but also incorporates a penalty for ties in model score which are not also paired with ties in wordlikeness ratings, or vis versa.

@@

A simple baseline is constructed by separating nonce words into those which contain a phonotactic violation and those which do not. As all nonce words here are monosyllabic, this task can be localized to two subcomponents of the syllable, the onset and the rime. This is not to imply that these are the only domains over which phonotactic violations might be stated, but there are prior claims that onset and rime are particularly important domains for stating phonotactic constraints (e.g., \citealt{Fudge1969}, \citealt{Kessler1997}, \citealt{Treiman2000}). Furthermore, these units are implicated by language games \citep{Treiman1983} and speech errors \citep{Fowler1993}, and adult speakers are adept at separating syllables into these units \citep{Treiman1986,Treiman1995}.

Operationalizing ``phonotactic violation'' is somewhat more difficult. The simplest possible mechanism is chosen here: an onset or rime is identified as well-formed if it occurs with non-zero frequency in a representative sample, and is identified as ill-formed otherwise. The sample is derived from those entries of the CMU pronunciation dictionary which occur at least once per million words in the SUBTLEX-US frequency norms, the latter though to be particualrly strongly correlated with behavioral measures \citep{Brysbaert2009}. These pronunciations are then syllabified, and individual syllables parsed into onset and rime, according to a process described in detail in Appendix \ref{syllabification}. This is not is not to imply that all unattested onsets or rimes should be regarded as ill-formed, or that all onsets or rimes with non-zero frequency in this data are well-formed. For instance, \citet{Albright2009a} judges [drEsp] to be an unobjectionable possible word of English, despite the total absense of [Esp] rimes in English, and similar observations have been made regarding English onsets \citep[e.g.,][]{Cairns1972,Moreton2002}.

@@

A second baseline is provided by measures of similarity to existing English words items, which has long been applied in the study of wordlikeness (e.g., \citealt{Greenberg1964}, \citealt{Ohala1986b}, \citealt{Shademan2006,Shademan2007}, \citealt{Vitevitch1998,Vitevitch1999a}). \citeauthor{LSLT} (\citeyear{LSLT}: 151, fn.~27) suggests that grammaticality judgements in general might be influenced by similarity to existing grammatical structures. \citet[417f.]{SPE} outline a similarity-based wordlikeness model. More recently, it has been observed (e.g., \citealt[51]{Coleman1997}, \citealt{Hay2004a}) that nonce words which flagrantly violate English sonority restrictions but which bear common affixes (e.g., *\emph{mrupation}) are rated highly English-like.

A wide variety of lexical similarity measures were considered, including a variant of including a variant of the Generalized Neighborhood model \citep{Bailey2001}, PLD20 \citep{Suarez2011}, and a set of measures provided by the Irvine Phonotactic Calculator \citep{Vaden2009}. The most reliable measure is also the most veneraable measure of lexical similarity: Coltheart's $N$ \citep{Coltheart1977}, which is defined as the number of words in some representative sample which can be changed into a target nonce word by a single insertion, deletion, or substitution of a phone. \citet{Greenberg1964} find a correlation between wordlikeness ratings and a variant of this measure which only counts words differing by a single substitution. This measure is plotted against ratings from the three studies in Figure \ref{neighborhood}, where it can be seen that it accounts for much of the variance in ratings.

@@

While there is nothing inherently ``phonotactic'' about Coltheart's $N$, it indirectly incorporates much of the information present in the gross phonotactic baseline. Consider [blIk]: since there is nothing marked about any part of this nonce word, a ``neighbor'' might be found by modifying any phone: e.g., \emph{click}, \emph{brick}, \emph{bloke}, \emph{bliss}. However, since [bn] onsets are unattested in English, a neighbor of [bnIk] must somehow modify this cluster: this leaves only \emph{brick} and \emph{nick}. \citet{Frauenfelder1993} find that neighborhood density is also strongly correlated with measures like bigram probability.

@@

Bigrams probabilities have been applied to model results of a number of psycholinguistic tasks \citep[e.g.,][]{Pitt1998,Vitevitch2005}, and \citet{Albright2009a} applies them as a model of wordlikeness judgements. The bigram probability of a sequence $ijk$, for instance, is:

>>>>

Recent developments in the theory of phonotactic knowledge are motivated by the claim that speakers' wordlikeness intuitions are inherently \emph{gradient}, i.e., that such intuitions are more granular than implied by a binary contrast between ``possible'' and ``impossible'' words.

\begin{quote}
When native speakers are asked to judge made-up (nonce) words, their intuitions are rarely all-or-nothing. 
In the usual case, novel items fall along a gradient cline of acceptability. \citep[][9]{Albright2009a}

In the particular domain of phonotactics gradient intuitions are pervasive: they have been found in every experiment that allowed participants to rate forms on a scale.
\citep[][382]{Hayes2008a}

\ldots{}when judgements are elicited in a controlled fashion from speakers, they always emerge as gradient, including all intermediate values. \citep[371]{Shademan2006} 
\end{quote}

This is not a novel claim. Early generative discussions of wordlikeness \citep[e.g.,][]{Chomsky1965,Halle1962} are best remembered for the famous examples [blɪk] and [bnɪk], the former representing a ``possible word'' of English and the latter representing an ``impossible word''. 
A naïve account of this contrast follows from the assumption that segments must be parsed into syllables or subject to further phonological repair (e.g., \citealt[10f.]{Hooper1973}, \citealt[57f.]{Kahn1976}, \citealt{Ito1989a}, \citealt[19f.]{Wolf2009}). 
Unlike some languages (e.g., Morroccan Arabic: \emph{bniqa} `closet'), English does not permit stop-nasal onsets like [bn], so the latter nonce word cannot surface as such. 
In other words, [bnɪk] is an impossible surface representation in English. 
However, in \emph{The Sound Pattern of English}, \citet{SPE} introduce a third nonce word, [bznk], and claim that last form is in some sense even less English-like than [bnɪk].\footnote{
    That wordlikeness judgements depend on language-specific knowledge is apparent given that [bznk] is not impossible in all languages: Imdlawn Tashlhiyt Berber permits whole words consisting of a stop-fricative-nasal-stop sequence (e.g., [tzmt] `it is stifling'; \citealt[112]{Dell1985}). 
    This suggests that wordlikeness judgements depend on language-specific knowledge.
    It has more recently been suggested some contrasts between unattested clusters can be viewed as cross-linguistic implicational universals \citep[e.g.,][]{Berent2007a,Berent2008a,Berent2009,Berent2007b}, so that, for instance, the acceptance of [bznk] might imply acceptance of [bnɪk]. 
    If this is correct, it undermines \citeauthor{SPE}'s position.}
From this they too conclude that wordlikeness intuitions are gradient.

\begin{quote}
Hence, a real solution to the problem of ``admissibility'' will not simply define a tripartite categorization of occurring, accidental gap, and inadmissible, but will define the `degree of admissibility' of each potential lexical matrix in such a way as to\ldots{}make numerous other distinctions of this sort (\emph{SPE}:416--417)
\end{quote}

\noindent
This brings the theory of wordlikeness in line with the view of syntactic grammaticality presented by foundational documents like \emph{The Logical Structure of Linguistic Theory} \citep{LSLT} and \emph{Aspects in the Theory of Syntax} \citep{ASPECTS}, which posits multiple degrees of ungrammaticality arising from different types of syntactic violations.
\citet{BARRIERS} and \citet{Huang1982} propose further elaborations of this type of theory (see \citealt[43f.]{Schutze1996} for a critique).

It has been recognized, however, that the naïve account of the [blɪk] and [bnɪk] cannot easily be extended to account for these ``numerous other distinctions''.

\begin{quote}
A defect of current grammatical accounts of phonotactics is that they render simple up-or-down decisions concerning well-formedness and cannot account for gradient judgements. \citep[371]{Shademan2006}
\end{quote}

Scientific observations do not necessarily arrive at the appropriate granularity for analysis.
Since the granularity of measurement and analysis is determined by prior hypotheses about the nature of the system under study, disputes are inevitable as scientific paradigms evolve. 
This study argues that there are both theoretical and empirical arguments to doubt the implicit hypothesis linking scalar wordlikeness judgements and a gradient model of phonotactics. First, intermediate judgements are characteristic of all gradient rating tasks, and therefore these judgements are irrelevant to the question of whether well-formedness is categorical or gradient.
Secondly, simple baselines better account for gradient well-formedness judgements than current computational models of phonotactic knowledge, suggesting that the gradience observed in these tasks do not derive from grammatical mechanisms.

%are ``\ldots{}under the obligation to say why \emph{what it takes to be} data relevant to the confirmation of its theories \emph{are} data relevant to the confirmation of its theories'' \citep[200]{Fodor1981b}.
%The naïve account of the [blɪk]/[bnɪk] contrast presented above is a special case of this

\section{Aspects in the theory of gradient grammaticality}

The aforementioned discussions of gradient aspects of wordlikeness judgements take for granted that intermediate judgements require a model of gradient grammaticality. 
The general position that such data can be taken at face value is known in the cognitive sciences as \emph{naïve realism} \citep{Fodor1981a}, and is not uncontroversial. 
With respect to wordlikeness, there are numerous reasons to question the wisdom of this position.

\subsection{What some linguistic intuitions might not be}

As first noted by \citet{Chomsky1963b}, speakers experience difficulty processing sentences with multiple center embeddings. \citet{Gibson1999} find that speakers rate sentences like (\ref{gibson}a), which is well-formed, less grammatical than (\ref{gibson}b), which is nonsensical.

\begin{example}[A well-formedness illusion]
\label{gibson}
\begin{tabular}{l l@{}l}
a. &    & The patient who the nurse who the clinic had hired admitted met Jack. \\
b. & *  & The patient who the nurse who the clinic had hired met Jack. \\
\end{tabular}
\end{example}

\noindent
It is informative to consider that this well-known result has had no effect on the theory of syntactic representations, only on the theory of linguistic memory; it is recognized as the product of cognitive restrictions found in non-linguistic domains, as a \emph{task effect}.
This contrasts with the argument made by \citet{Hayes2000}, that gradient wordlikeness judgements demand an all-encompassing revision to the grammatical architecture, reviewed below.

The results of controlled experiments are often biased by subtle details that seem orthogonal to the task: for instance, certain types of duration judgements are systematically biased by consumption of caffeine \citep{Gruber2005}. 
It should come as no surprise, then, that a highly salient aspect of a judgement task, the scale used for responses, also influences the results obtained.
\citet*{Armstrong1983} argue that the use of many-valued scales may cause intermediate ratings, and should be interpreted as a task effect.

\citet{Armstrong1983} are concerned with experimental evidence for the nature of cognitive concepts. 
While they do not attempt to dispute that certain concepts (e.g., \emph{fruit}) have a family-resemblance structure \citep[e.g.,][]{Rosch1975a}, they assert that it is apparent that other concepts are ``definitional'' (i.e., all-or-nothing), a notion which they illustrate with \emph{odd number}.

\begin{quote}
No integer seems to sit on the fence, undecided as to whether it is quite even, or perhaps a bit odd. No odd number seems odder than any other odd number. \citep[274]{Armstrong1983}
\end{quote}

\noindent
However, when subjects are asked to rate, using a 7-point Likert scale, how representative individual odd counting numbers are of the concept \emph{odd number}, they freely use intermediate ratings; their results with \emph{odd number} and \emph{even number} are shown in Figure \ref{agg}.

@@

This suggests that the gradience observed is primarily an artifact of the task itself. \citeauthor{Schutze2011} suggests that the nature of this effect might be understood as the result of speakers' attempts to reconcile bizarre experimental tasks with their knowledge.

\begin{quote}
Putting it another way, when asked for gradient responses, participants will find some way to oblige the experimenter; if doing so is incompatible with the experimenter's actual question, they apparently infer that she must have really intended to ask something slightly different. \citep[24]{Schutze2011}
\end{quote}

As \citeauthor{Armstrong1983} observe, these results show that the scalar judgement tasks provide no evidence as to whether the category being rated is categorical or gradient.

\begin{quote}
\ldots{}we hold that \emph{fruit} and \emph{odd number} have different structures, and yet we obtain the same experimental outcome for both. But if the same result is achieved regardless of the concept structure, then the experimental design is not pertinent to the determination of concept structure. \citep[284--5]{Armstrong1983}
\end{quote}

It might be said that these results reveal something about the representation of odd numbers, but no scientist of mathematical cognition has risen to the challenge of constructing a theory that might account for the fact that 447 is rated more odd than 3. 
\citeauthor{Armstrong1983} anticipate this objection.

\begin{quote}
Some have responded to these findings very consistently, by asserting that the experimental findings are to be interpreted as before: that, psychologically speaking, odd numbers as well as birds and vegetables are graded concepts\ldots{} We reject this conclusion just because we could not explain how a person could compute with integers who believed that 7 was odder than 23. We assert confidently that the facts about subjects being able to compute and about their being able to give the definition of odd number, etc., are the more important, highly entrenched, facts we want to preserve and explain\ldots{} \citep[284]{Armstrong1983}
%we ourselves are prepared to give up the seeming fact that some odd numbers appear, as shown by their behavior in certain experimental paradigms, to be odder than others\ldots{}we do not give it up by saying that it was no fact; rather, by saying it must have been a fact about something other than the structure of concepts. \citep[284]{Armstrong1983}
\end{quote}

\subsection{A model of gradient intuitions}

It is possible to draw an analogy: can it be the case that [st] is a significantly ``better'' onset than [bl] (a prediction of the wordlikeness model proposed by \citealt{Albright2009a}, for instance), but ensure that both are treated the same with respect to syllabification? 
Current research on gradient well-formedness is concerned with specifying the component by which a scalar value is assigned to linguistic structures. 
This is only one part of any model of gradient grammaticality, however.
Further assumptions are necessary, as follows.
When presented with a linguistic item (of whatever type) in a well-formedness task, the grammar assigns a parse.
This grammar must be able to parse an enormous range of linguistic structures, including many which it cannot be permitted to generate; independent perception and production grammars may be necessary. 
A scalar value is then assigned to this parse. 
Then, speakers consciously access this scalar value, then transform it in accordance with the numerical scale chosen by the experimenter.

Each step of this procedure is dubious, however.
First, speakers have difficulty perceiving \citep{Brown1956} and producing \citep{Davidson2005,Davidson2006a,Davidson2006b,Davidson2010,GallagherInPress,Rose2007,Vitevitch1998,Vitevitch2005} phonotactically illicit non-words, suggesting that speakers' ability to faithfully parse illicit representations is at best quite limited.
Secondly, the computation of a scalar value serves no further purpose than to provide for gradient well-formedness judgements; even if this is the result of probabilistic parsing \citep[e.g.,][]{Coleman1997}, 
it need not be computed overtly.\footnote{
    An objection might be made here on evolutionary grounds. 
    It is still quite mysterious why the human language endowment includes constraints on pronominal binding, for instance, but these constraints are implicated in everyday language use.
    Far more bizarre is the suggestion that the linguistic endowment includes mechanisms only used in certain experimental tasks.}
Next, speakers must be able to consciously access and report the magnitude of this value (it must be \emph{cognitively penetrable} in the sense of \citealt{Pylyshyn1984}), an ability which is limited in many other domains.
Finally, \citet{Sprouse2011} argues that speakers do not (or cannot) scale well-formedness values to satisfy the assumptions magnitude estimation, a type of gradient rating task.
%\citep{Bard1996}

It is informative to compare this baroque architecture to a model of what is necessary to make a binary well-formedness judgement. When presented with a linguistic item in a judgement task, the grammar attempts to assign a parse. Speakers then access whether or not parsing was successful. 
There are reasons to think that parsing of ungrammatical structures does in fact result in a ``crash'': whereas syntactic priming increases the acceptability of grammatical structures \citep{Luka2005}, ungrammatical structures show no priming effects \citep{Sprouse2007b}. 
As priming of linguistic structures is thought to implicate shared representations in memory, ungrammatical structures may not stored in linguistic memory. 
%This cannot be reduced to the fact that ungrammatical structures fail to denote, since well-formed but non-existent structures do produce facilitory priming \citep[e.g.,][]{Longtin2005}. 
The fact that requests for repetition and clarification are ubiquituous in spontaneous speech illustrates further that speakers are frequently aware when parsing has failed.
Consequently, a great deal of evidence is needed to reject this simple model in favor of the gradient grammaticality architecture.

\subsection{Evidencing gradience}

\citet{Hayes2000} argues that it is ``uninsightful'' to attribute gradience to task effects, insofar as these effects call upon grammatical representations.

%\citep[?]{Hayes2000}
%``much of the patterning of gradient judgments is based on authentic structural aspects of the linguistic material being judged''

\begin{quote}
\ldots{}patterns of gradient well-formedness often seem to be driven by the very same principles that govern absolute well-formedness\ldots{} I conclude that the proposed attribution of gradient well-formedness judgments to performance mechanisms would be uninsightful. Whatever ``performance'' mechanisms we adopted would look startlingly like the grammatical mechanisms that account for non-gradient judgments. \citep[99]{Hayes2000}
\end{quote}

\noindent
This is indisputable. 
However, there have been no serious attempts to evaluate categorical and gradient models of wordlikeness on an equal footing.
In light of the complexities of gradient models, such an evaluation requires strong quantitative evidence for the superiority of gradient grammatical models.
This study represents a first attempt to fill this gap.

It is not that categorical models have been ignored by the literature on wordlikeness modeling, but rather that they have not been compared. 
\citet{Frisch2000} and \citet{Vitevitch1997} find that speakers' wordlikeness ratings of multisyllabic words are correlated with a probabilistic measure of the well-formedness of the constituent syllables. 
Unfortunately, no attempt is made to control for the well-formedness of syllable contact clusters in these words: some of the stimuli have medial consonant clusters containing both voiced and voiceless obstruents (e.g., [ɡaɪbsaɪk]), something which is exceptionally rare in English simplex words.
Similarly, \citet{Hayes2008a}, who compare their gradient model of wordlikeness against a set of phonotactic constraints proposed by \citet{Clements1983}, first transform these constraints, many of which are exceptionless, into probabilities. 
While this is consistent with their claim, that ``the ability to model gradient intuitions to be an important criterion for evaluating phonotactic models'' \citep[382]{Hayes2008a}, this principle would preclude any attempt to test the hypotheses that underlies it.

%(see \citealt{Sproat1993} and \citealt{Wrench2003} for articulatory investigations).

@@

It is unknown whether the intermediate ratings in gradient wordlikeness tasks are reliably predicted by computational models that have been proposed. 
If some model poorly accounts for these intermediate ratings, it may be the case that the model is improperly specified, or it may indicate that a considerable amount of variance in wordlikeness ratings is not a product of the grammatical system per se. 
It seems particularly plausible that speakers might differentiate, in a regular fashion, between different types of ``impossible'' words, as proposed by \citet{SPE}. 
There are also claims that speakers distinguish between different types of ``possible'' words, so that, for instance, [stɪn] \emph{stin} is rated more English-like than [blɪn] \emph{blin} \citep{Albright2009a}. 
Alternatively, wordlikeness judgements could be effectively modeled with a gross contrast between possible and impossible words; even then, a gradient model might show a residual correlation.
All of these possibilities are considered here.

@@

This evaluation uses a large sample of three previously published studies on English wordlikeness comprising 125 subjects and 187 items. Two criteria were used to select these three studies. First, it was required that the stimuli be presented aurally so as to eliminate any possibility of orthographic effects \citep[e.g.,][]{Berent2001b,Berent2008b}. Secondly, the data must be sufficiently ``phonotactically diverse'': that is, it must include both items like \emph{blick} and \emph{bnick}. This excludes studies like that of \citet{Bailey2001}, in which few if any items contain gross phonotactic violations of the type represented by \emph{bnick}, since this is likely to minimize the coverage of any grammatical model. The data is summarized in Table \ref{counts}.

@@

\citet{Albright2007} administers a wordlikeness task in which 68 adult speakers rate 40 monosyllabic nonce words, presented aurally, on a 7-point Likert scale with endpoints labeled  ``completely impossible as an English word'' and ``would make a fine English word''. \citeauthor{Albright2007}'s study is primarily concerned with the effects of different onset types (e.g., well-formed /bl/, marginal /bw/, unattested /bn, bd, bz/), and there is less diversity among the choice of rimes, none of which are obviously ill-formed.

@@

\citet{Scholes1966} conducts several wordlikeness tasks with 7th-grade students (approximately 12--13 years of age). The data used here is his experiment 5, in which 33 speakers provide a ``yes'' or ``no'' as to whether each of the 63 items, presented aurally, are ``likely to be usable as a word of English''. Like the study by \citet{Albright2007}, the focus is on onset well-formedness and there is minimal diversity in rime type. Two items, [klʌŋ] \emph{clung} and [bɹʌŋ] \emph{brung} (a dialectical past participle of \emph{bring}), are excluded here as actual words of English. \citet{Albright2009a} and \citet{Hayes2008a} also use this data for the purposes of model evaluation; following \citet{Frisch2000}, they use the proportion of ``yes'' responses for each item so as to derive a continuous measure of well-formedness.

@@

Models are evaluated by comparing their scores to the average rating of each word using four correlation statistics. Each of these range between $[-1, 1]$, where $1$ indicates a perfect positive correlation and $-1$ denotes a perfect negative correlation. \citet{Hayes2008a} evaluate their model using the Pearson (``product-moment'') $r$, a parametric correlation measure.  \citet{Stevens1946} argues that statistics of this type are inappropriate for analysis of Likert scale data, like those used by \citet{Albright2007} and \citet{Albright2003b}.
The reason is that the Pearson $r$ makes a \emph{linearity assumption}. That is, it assumes that nonce words rated ``1'' and  ``3'', for instance, are just as different as those ``4'' and a ``6''. A weaker assumption, more appropriate for Likert scale data, is the \emph{monotonicity assumption}: that ``1'' is less English-like than ``3'', which is less English-like than ``4'', and so on. However, it also has been claimed that $r$ is particularly robust to violations of the linearity assumption \citep[e.g.,][]{Havlicek1976}. Pearson $r$ is reported here, but no stance is taken on its appropriateness for this data.
%\citet{Jamieson2004}
\citeauthor{Hayes2008a} also report Spearman $\rho$; this statistic requires only the weaker assumption of monotonicity, but it is difficult to give a simple interpretation to the coefficient.

Much easier to interpret are two other non-parametric statistics, the Goodman-Kruskal $\gamma$ and the Kendall $\tau_b$ \citep{Noether1981}, as follows. These statistics are computed by comparing every model score/wordlikeness rating pair to every other: a comparison is counted as \emph{concordant} if the greater of the two model scores is the one associated with the greater of the two wordlikeness ratings (that is, the model ranks these two nonce words the in accordance with speakers' ratings), and as \emph{discordant} otherwise. These two statistics differ only in the treatment of ``ties'', pairs where either the model score or wordlikeness rating are identical. For $\gamma$, ties are ignored, and the coefficient is

\begin{equation*}
\gamma = \displaystyle\frac{c - d}{c + d}
\end{equation*}

\noindent
where $c$ and $d$ represent the number of concordant and discordant pairs, respectively. The $\tau_b$ statistic uses a similar formula, but also incorporates a penalty for ties in model score which are not also paired with ties in wordlikeness ratings, or vis versa. \citet{Albright2009a} uses a variant of this statistic to evaluate wordlikeness models.

@@

A simple baseline is constructed by separating nonce words into those which contain a phonotactic violation and those which do not. As all nonce words here are monosyllabic, this task can be localized to two subcomponents of the syllable, the onset and the rime. This is not to imply that these are the only domains over which phonotactic violations might be stated, but there are prior claims that onset and rime are particularly important domains for stating phonotactic constraints (e.g., \citealt{Fudge1969}, \citealt{Kessler1997}, \citealt{Treiman2000}). Speakers are adept at separating syllables into these units \citep{Treiman1983,Treiman1986,Treiman1995}, and they are implicated by patterns of speech errors \citep{Fowler1987,Fowler1993}.

Operationalizing ``phonotactic violation'' is somewhat more difficult. The simplest possible mechanism is chosen here: an onset or rime is identified as well-formed if it occurs with non-zero frequency in a representative sample, and is identified as ill-formed otherwise. The sample is derived from those entries of the CMU pronunciation dictionary which occur at least once per million words in the SUBTLEX-US frequency norms, the latter though to be particularly strongly correlated with behavioral measures \citep{Brysbaert2009}. These pronunciations are then syllabified, and individual syllables parsed into onset and rime, according to a process described in detail in Appendix \ref{syllabification}. This is not is not to imply that all unattested onsets or rimes should be regarded as ill-formed, or that all onsets or rimes with non-zero frequency in this data are well-formed. For instance, \citet{Albright2009a} judges [dɹɛsp] \emph{dresp} to be phonotactically well-formed, despite the total lack of [ɛsp] rimes in English. Similar observations have been made concerning English onsets \citep[e.g.,][]{Cairns1972,Moreton2002}. In the other direction, there are precedents for labeling certain attested words as phonotactically ``peripheral'' (e.g., \citealt{Myers1987}, \citealt{Borowsky1989}), as lexical exceptions to language-specific prosodic constraints, without labeling these words as absolutely ungrammatical. Neither of these dissociations between attestation and well-formedness are implemented here, however.

@@

A second baseline is provided by measures of similarity to existing English words items, which has long been applied to model wordlikeness judgements (e.g., \citealt{Bailey2001}, \citealt{Greenberg1964}, \citealt{Kirby2007}, \citealt{Ohala1986b}, \citealt{Shademan2006,Shademan2007}, \citealt{Vitevitch1998,Vitevitch1999a}). \citeauthor{LSLT} (\citeyear{LSLT}: 151, fn.~27) suggests that grammaticality judgements in general might be influenced by similarity to existing grammatical structures, and \citet[417f.]{SPE} outline a similarity-based wordlikeness model. More recently, it has been observed (e.g., \citealt[51]{Coleman1997}, \citealt{Hay2004a}) that nonce words which flagrantly violate English sonority restrictions but which bear common affixes (e.g., *\emph{mrupation}) are rated highly English-like.

A wide variety of lexical similarity measures were considered, including a variant of including a variant of the Generalized Neighborhood model \citep{Bailey2001}, PLD20 \citep{Suarez2011}, and a set of measures provided by the Irvine Phonotactic Calculator \citep{Vaden2009}. The most reliable measure is also the most venerable measure of lexical similarity: Coltheart's $N$ \citep{Coltheart1977}, which is defined as the number of words in some representative sample which can be changed into a target nonce word by a single insertion, deletion, or substitution of a phone. \citet{Greenberg1964} find a correlation between wordlikeness ratings and a variant of this measure which only counts words differing by a single substitution. This measure is plotted against ratings from the three studies in Figure \ref{neighborhood}, where it can be seen that it accounts for much of the variance in ratings.

%The effects of neighborhood density in various psycholinguistic tasks are emergent properties of many models of speech production \citep[e.g.,][]{Luce1998,Luce2000} and perception \citep{Marslen-Wilson1984,Marslen-Wilson1987,McClelland1986,Norris1994,Norris2000}.

@@

While there is nothing inherently ``phonotactic'' about Coltheart's $N$, it indirectly incorporates much of the information present in the gross phonotactic baseline. 
Consider [blɪk]: since there is nothing marked about any part of this nonce word, a ``neighbor'' might be found by modifying any phone: e.g., \emph{click}, \emph{brick}, \emph{bloke}, \emph{bliss}. 
However, since [bn] onsets are unattested in English, a neighbor of [bnɪk] must somehow modify this cluster: this leaves only \emph{brick} and \emph{nick}. 
\citet{Bailey2001} and \citet{Frauenfelder1993} note that neighborhood density is also strongly correlated with measures like bigram probability, but it is argued elsewhere that phonotactic measures and neighborhood density have distinct effects \citep[e.g.,][]{Berent2003,Pitt1998b,Vitevitch1998,Vitevitch1999a}.

@@

Faciliatory effects of bigram probabilities (i.e., shorter latencies) are reported for other nonce word tasks conducted with adults, including single-word shadowing \citep{Vitevitch1997,Vitevitch1998}, same/different judgements \citep{Lipinski2005,Luce2001,Vitevitch1999a,Vitevitch2005}, and lexical decision \citep{Pylkkanen2002a}. \citet{Albright2009a} applies them as a model of wordlikeness judgements. The bigram probability of a sequence $ijk$, for instance, is:
